Impact of Satellite Constellations on the Legacy Survey of Space and Time

### 2022-09-28 ms.tex
draft 1

%% Beginning of file 'sample631.tex'
%%
%% Modified 2022 May  
%%
%% This is a sample manuscript marked up using the
%% AASTeX v6.31 LaTeX 2e macros.
%%
%% AASTeX is now based on Alexey Vikhlinin's emulateapj.cls 
%% (Copyright 2000-2015).  See the classfile for details.

%% AASTeX requires revtex4-1.cls and other external packages such as
%% latexsym, graphicx, amssymb, longtable, and epsf.  Note that as of 
%% Oct 2020, APS now uses revtex4.2e for its journals but remember that 
%% AASTeX v6+ still uses v4.1. All of these external packages should 
%% already be present in the modern TeX distributions but not always.
%% For example, revtex4.1 seems to be missing in the linux version of
%% TexLive 2020. One should be able to get all packages from www.ctan.org.
%% In particular, revtex v4.1 can be found at 
%% https://www.ctan.org/pkg/revtex4-1.

%% The first piece of markup in an AASTeX v6.x document is the \documentclass
%% command. LaTeX will ignore any data that comes before this command. The 
%% documentclass can take an optional argument to modify the output style.
%% The command below calls the preprint style which will produce a tightly 
%% typeset, one-column, single-spaced document.  It is the default and thus
%% does not need to be explicitly stated.
%%
%% using aastex version 6.3
\documentclass[linenumbers]{aastex631}

%% The default is a single spaced, 10 point font, single spaced article.
%% There are 5 other style options available via an optional argument. They
%% can be invoked like this:
%%
%% \documentclass[arguments]{aastex631}
%% 
%% where the layout options are:
%%
%%  twocolumn   : two text columns, 10 point font, single spaced article.
%%                This is the most compact and represent the final published
%%                derived PDF copy of the accepted manuscript from the publisher
%%  manuscript  : one text column, 12 point font, double spaced article.
%%  preprint    : one text column, 12 point font, single spaced article.  
%%  preprint2   : two text columns, 12 point font, single spaced article.
%%  modern      : a stylish, single text column, 12 point font, article with
%% 		  wider left and right margins. This uses the Daniel
%% 		  Foreman-Mackey and David Hogg design.
%%  RNAAS       : Supresses an abstract. Originally for RNAAS manuscripts 
%%                but now that abstracts are required this is obsolete for
%%                AAS Journals. Authors might need it for other reasons. DO NOT
%%                use \begin{abstract} and \end{abstract} with this style.
%%
%% Note that you can submit to the AAS Journals in any of these 6 styles.
%%
%% There are other optional arguments one can invoke to allow other stylistic
%% actions. The available options are:
%%
%%   astrosymb    : Loads Astrosymb font and define \astrocommands. 
%%   tighten      : Makes baselineskip slightly smaller, only works with 
%%                  the twocolumn substyle.
%%   times        : uses times font instead of the default
%%   linenumbers  : turn on lineno package.
%%   trackchanges : required to see the revision mark up and print its output
%%   longauthor   : Do not use the more compressed footnote style (default) for 
%%                  the author/collaboration/affiliations. Instead print all
%%                  affiliation information after each name. Creates a much 
%%                  longer author list but may be desirable for short 
%%                  author papers.
%% twocolappendix : make 2 column appendix.
%%   anonymous    : Do not show the authors, affiliations and acknowledgments 
%%                  for dual anonymous review.
%%
%% these can be used in any combination, e.g.
%%
%% \documentclass[twocolumn,linenumbers,trackchanges]{aastex631}
%%
%% AASTeX v6.* now includes \hyperref support. While we have built in specific
%% defaults into the classfile you can manually override them with the
%% \hypersetup command. For example,
%%
%% \hypersetup{linkcolor=red,citecolor=green,filecolor=cyan,urlcolor=magenta}
%%
%% will change the color of the internal links to red, the links to the
%% bibliography to green, the file links to cyan, and the external links to
%% magenta. Additional information on \hyperref options can be found here:
%% https://www.tug.org/applications/hyperref/manual.html#x1-40003
%%
%% Note that in v6.3 "bookmarks" has been changed to "true" in hyperref
%% to improve the accessibility of the compiled pdf file.
%%
%% If you want to create your own macros, you can do so
%% using \newcommand. Your macros should appear before
%% the \begin{document} command.
%%
\newcommand{\vdag}{(v)^\dagger}
\newcommand\aastex{AAS\TeX}
\newcommand\latex{La\TeX}

%% Reintroduced the \received and \accepted commands from AASTeX v5.2
%\received{March 1, 2021}
%\revised{April 1, 2021}
%\accepted{\today}

%% Command to document which AAS Journal the manuscript was submitted to.
%% Adds "Submitted to " the argument.
%\submitjournal{PSJ}

%% For manuscript that include authors in collaborations, AASTeX v6.31
%% builds on the \collaboration command to allow greater freedom to 
%% keep the traditional author+affiliation information but only show
%% subsets. The \collaboration command now must appear AFTER the group
%% of authors in the collaboration and it takes TWO arguments. The last
%% is still the collaboration identifier. The text given in this
%% argument is what will be shown in the manuscript. The first argument
%% is the number of author above the \collaboration command to show with
%% the collaboration text. If there are authors that are not part of any
%% collaboration the \nocollaboration command is used. This command takes
%% one argument which is also the number of authors above to show. A
%% dashed line is shown to indicate no collaboration. This example manuscript
%% shows how these commands work to display specific set of authors 
%% on the front page.
%%
%% For manuscript without any need to use \collaboration the 
%% \AuthorCollaborationLimit command from v6.2 can still be used to 
%% show a subset of authors.
%
%\AuthorCollaborationLimit=2
%
%% will only show Schwarz & Muench on the front page of the manuscript
%% (assuming the \collaboration and \nocollaboration commands are
%% commented out).
%%
%% Note that all of the author will be shown in the published article.
%% This feature is meant to be used prior to acceptance to make the
%% front end of a long author article more manageable. Please do not use
%% this functionality for manuscripts with less than 20 authors. Conversely,
%% please do use this when the number of authors exceeds 40.
%%
%% Use \allauthors at the manuscript end to show the full author list.
%% This command should only be used with \AuthorCollaborationLimit is used.

@@

\title{Impact of Mega Satellite Constellations on the Legacy Survey of Space and Time}

\author[0000-0003-2874-6464]{Peter Yoachim}
\affiliation{Department of Astronomy, University of Washington, Seattle, WA, USA}

XXX--add Meredith, Alina, Kresimir, Zeljko
\author{more authors}
\affiliation{more affilt}

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}

abstract
\end{abstract}

@@

The overview paper \citep{overview}

\section{Constellations}

>>>>

%% using aastex version 6.3
\documentclass[linenumbers]{aastex631}

%% The default is a single spaced, 10 point font, single spaced article.
%% There are 5 other style options available via an optional argument. They
%% can be invoked like this:
%%
%% \documentclass[arguments]{aastex631}
%% 
%% where the layout options are:
%%
%%  twocolumn   : two text columns, 10 point font, single spaced article.
%%                This is the most compact and represent the final published
%%                derived PDF copy of the accepted manuscript from the publisher
%%  manuscript  : one text column, 12 point font, double spaced article.
%%  preprint    : one text column, 12 point font, single spaced article.  
%%  preprint2   : two text columns, 12 point font, single spaced article.
%%  modern      : a stylish, single text column, 12 point font, article with
%% 		          wider left and right margins. This uses the Daniel
%% 		          Foreman-Mackey and David Hogg design.
%%  RNAAS       : Suppresses an abstract. Originally for RNAAS manuscripts 
%%                but now that abstracts are required this is obsolete for
%%                AAS Journals. Authors might need it for other reasons. DO NOT
%%                use \begin{abstract} and \end{abstract} with this style.
%%
%% Note that you can submit to the AAS Journals in any of these 6 styles.
%%
%% There are other optional arguments one can invoke to allow other stylistic
%% actions. The available options are:
%%
%%   astrosymb    : Loads Astrosymb font and define \astrocommands. 
%%   tighten      : Makes baselineskip slightly smaller, only works with 
%%                  the twocolumn substyle.
%%   times        : uses times font instead of the default
%%   linenumbers  : turn on lineno package.
%%   trackchanges : required to see the revision mark up and print its output
%%   longauthor   : Do not use the more compressed footnote style (default) for 
%%                  the author/collaboration/affiliations. Instead print all
%%                  affiliation information after each name. Creates a much 
%%                  longer author list but may be desirable for short 
%%                  author papers.
%% twocolappendix : make 2 column appendix.
%%   anonymous    : Do not show the authors, affiliations and acknowledgments 
%%                  for dual anonymous review.
%%
%% these can be used in any combination, e.g.
%%
%% \documentclass[twocolumn,linenumbers,trackchanges]{aastex631}
%%
%%
%% If you want to create your own macros, you can do so
%% using \newcommand. Your macros should appear before
%% the \begin{document} command.
%%
\newcommand{\vdag}{(v)^\dagger}
\newcommand\aastex{AAS\TeX}
\newcommand\latex{La\TeX}

%\received{March 1, 2021}
%\revised{April 1, 2021}
%\accepted{\today}

%% Command to document which AAS Journal the manuscript was submitted to.
%% Adds "Submitted to " the argument.
%\submitjournal{ApJL}

%%
%% Note that all of the author will be shown in the published article.
%% This feature is meant to be used prior to acceptance to make the
%% front end of a long author article more manageable. Please do not use
%% this functionality for manuscripts with less than 20 authors. Conversely,
%% please do use this when the number of authors exceeds 40.
%%
%% Use \allauthors at the manuscript end to show the full author list.
%% This command should only be used with \AuthorCollaborationLimit is used.

@@

\title{Impact of Satellite Constellations on the Legacy Survey of Space and Time}

\author[0000-0002-8400-1910]{Jinghan Alina Hu}
\affiliation{Harvey Mudd College, Claremont, CA, USA}
\author[0000-0003-1305-7308]{Meredith L. Rawls}
\affiliation{Department of Astronomy and DiRAC, University of Washington, Seattle, WA, USA}
\author[0000-0003-2874-6464]{Peter Yoachim}
\affiliation{Department of Astronomy and DiRAC, University of Washington, Seattle, WA, USA}
\author{Zeljko Ivezic}
\affiliation{Department of Astronomy and DiRAC, University of Washington, Seattle, WA, USA}
\author{Kresimir Mihic}

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}
Best abstract ever

\end{abstract}

@@

Vera C. Rubin Observatory's Legacy Survey of Space and Time (LSST) is a ten-year
astronomical imaging survey that will begin in 2024 from a new telescope in Chile. Instead
of soliciting individual requests for what the telescope should point at, the LSST will uniformly
survey the sky every three nights in six color filters to essentially create a decade-long
high-resolution survey of the entire southern sky, and share massive quantities of data products
with the astronomy community \citep{overview}.

To accomplish this, there is a complex scheduler algorithm that takes into account various
science priorities, including slew time, survey uniformity, and image depth \citep{naghib19}. The Rubin
Scheduler algorithm creates weightings for various science priorities and uses the weighting to
create a path for pointing observation.

However, one challenge for the LSST is that increasing numbers of bright
Low-Earth-Orbit (LEO) satellites (e.g., Starlink) are being launched, which could leave streaks
in astronomical pointings. Over the last three years, many astronomers have raised concerns
about the impact of this on the LEO ecosystem and astronomical surveys \citep{lawrence22,tyson20}. LEO satellites
are visible from Earth because they reflect sunlight, especially during twilight. As the
illuminated LEO satellites move across the field of view of an astronomical pointing, they leave
a streak in the image, which negatively impacts the scientific value of the pointing. Astronomers
predict there will be at least one satellite streaks in the majority of exposures of the Rubin
Observatory \citep{lawrence22}.An example of satellite streaks in an astronomical image is shown in Figure \ref{fig-streaks}.

\begin{figure}[ht!]\label{fig-streaks}
%\includegraphics{}  % preferably .eps
\caption{This is a caption}
\end{figure}

There have been efforts to reduce the impact of satellite streaks in astronomical pointings.
Satellite companies like SpaceX have worked on darkening the exterior of satellites so they will
be less illuminated [CITE SPACEX BLOG POST]. However, even with extra darkening, the Starlink satellites are still one
magnitude too bright for the observatory cameras, meaning these darkened satellite movements
will still be captured by the camera. Astronomers have worked on algorithms for masking
satellite trails in images, but covering the outer rim of the trails without losing extra pixels
remains a challenge \citep{tyson20}.
The rapid increase in population of LEO satellites threatens to compromise the quality
and scientific value of the LSST images and also necessitates expensive computational power to
mask the trails. Thus, we explore another option: incorporating known commercial satellites into
the Rubin Scheduler so the worst of them may be avoided. In this project, we create realistic
simulated forecasts of satellite trajectories and brightness, build a tool that uses that data to create
new scheduler constraints, and test the impact of the new scheduler on various Rubin LSST
observing programs.

%\section{Constellations}

\section{Methodology}
We begin by creating realistic forecasts of three different commercial satellite constellations: StarlinkV1 (4408 satellites, altitude 550 km), OneWeb (6372 satellites, altitude 550km), and StarlinkV2 (29988 satellites, altitude 550km). All three of these constellations are highly likely to launch as designed, or close to it, as both operators presently have functional satellites in orbit and are working to build and launch more. These simulated satellite
constellations are illustrated in Figure \ref{fig-simulated-constellations}.

\begin{figure}[ht!]\label{fig-simulated-constellations}
%\includegraphics{}  % preferably .eps
\caption{This is a caption}
\end{figure}

Using these simulated satellite constellations, we add an additional weighting for satellite
dodging to the Rubin Scheduler. Given an input satellite distribution, the scheduler predicts
satellite maps across a determined time frame. Based on the predicted location of satellites, the
scheduler creates a reward weighting: areas with more illuminated satellites will receive a more
negative reward weighting, or in other words, the scheduler is incentivized to avoid them. The
scheduler then combines the satellite reward weights with other weights to determine where to
point the telescope to take the next set of simulated observations. Figure \ref{fig-simulated-scheduler} shows an example of
scheduler weighting for the three simulated satellite constellations at two different sun altitudes.

\begin{figure}[ht!]\label{fig-simulated-scheduler}
%\includegraphics{}  % preferably .eps
\caption{This is a caption}
\end{figure}

As shown in Figure \ref{fig-simulated-scheduler}, the predicted OneWeb constellation map has more negative reward
weighting than other constellations. Although the OneWeb constellation has fewer satellites than
the StarlinkV2 constellation, the OneWeb constellation is at a higher altitude, meaning that its
satellites will be illuminated for a longer portion of the night, and not just close to twilight.
Therefore, OneWeb shows more negative weighting than StarlinkV2 though it has a smaller
satellite population.
To investigate whether the scheduler behaves how we expect with the new satellite
weighting added we create a testing function that measures the length of satellite streaks in the
simulated field of views. To ensure efficiency, only satellites that are above the altitude limit and
illuminated by the sun will be considered. Satellites below the altitude limit (indicated in the gray
region in Figure \ref{fig-simulated-scheduler}) can’t be observed in pointings and are therefore not included for efficiency.
For each satellite, we first determine whether they are in the field of view for any given
telescope pointing by calculating their distance from the center of the field of view. If their
distance from the field of view is less than the radius of the field of view, the satellite has crossed
through the pointing. To quantify the impact of the satellite on the pointing, we then project the satellite location as well as the pointing to a 2D x,y plane. In the 2D plane, the field of pointing
becomes roughly a circle and the start and end locations of the satellite become two points on the
plane. Using the shapely python library, we calculate the intersection length of the satellite location
points and the circle.
Therefore, with a given simulated satellite constellation and a schedule of observations,
we record the number of satellites that occur in each field of view and measure how long the
satellite streaks are. Using a realistic estimated satellite streak width of 50 pixels, we can convert
the streak length results into the number of affected image pixels, which allows us to
quantitatively measure the efficiency of the dodging algorithm with pixel loss.

\section{Results}
We begin by investigating the efficiency of the added weighting algorithm on dodging
satellites in observations. As shown in Figure \ref{fig-pixel-loss-weight}, using a conservative estimate of 50 pixels as
streak width, we found that higher dodging weights reduces pixels lost to satellite streaks. We
also found that smaller constellations at lower orbital altitudes (StarlinkV1 for example)
inherently results in less pixel loss per pointing, nearly independent of the dodging weight.
Figure \ref{fig-pixel-loss-weight} shows that the added weighted dodging algorithm was able to effectively avoid satellite streaks in pointings.

\begin{figure}[ht!]\label{fig-pixel-loss-weight}
%
%\includegraphics{}  % preferably .eps
\caption{This is a caption}
\end{figure}

We then investigate the relationship between number of exposures taken and the dodging
weight. As shown in Figure \ref{fig-pixel-loss-weight}, higher dodging weight results in fewer exposures, most likely due
to longer slew times. With the additional avoidance weighting, the telescope may be prompted to
slew to a location other than the desired pointing location and then slew back in order to avoid
some satellites, resulting in fewer overall exposures. We also found that larger constellations like
StarlinkV2 tend to decrease the number of exposures slightly more than smaller constellations
(StarlinkV1), which is expected. More satellites or satellites at higher orbital altitudes result in
bigger areas of avoidance on the sky, which results in more slewing to avoid the affected areas,
which subsequently reduces exposures. \\

We now investigate the tradeoff between survey depth and satellite avoidance. One goal of LSST is to collect a large number of exposures of all parts of the southern sky so the
exposures can be coadded to reveal faint structures that aren’t visible in individual images.
Therefore, survey depth is crucial to LSST science goals and the trade off must be evaluated.
With the satellite avoidance algorithm, the scheduler is prompted to avoid regions with
illuminated satellites, which could result in longer slew time or less desirable pointing
conditions, which contribute to a loss in survey depth. Therefore, we evaluate the tradeoff
between survey depth loss and satellite avoidance using the dodging algorithm.\\

\begin{figure}[ht!]\label{fig-trade-off}
%\includegraphics{}  % preferably .eps
\caption{This is a caption}
\end{figure}

As shown in Figure \ref{fig-trade-off} , for all three constellations, minor reduction in mean pixel loss per visit
($0.001\%$ decrease) results in a significant decrease in median coadded depth. Currently, pixel
loss due to satellite streaks only constitutes a very low percentage of total pixels in a pointing
(about $0.008\%$ without any dodging). Given that the percentage is currently so small, a very
minor reduction in mean pixel loss per visit results in significant reduction in image depth.

\section{Discussion and Conclusion}

We have demonstrated that adding a weighted term in the scheduler algorithm for
illuminated satellites can effectively reduce the amount of satellite streaks in observations, and
subsequently reduce mean pixel loss per visit. This is shown in Figure \ref{fig-pixel-loss-weight} where all three
constellations caused less pixel loss with increasing dodging weights. However, with the new
added priority on dodging, the telescope can be pushed to take an observation path that does not
optimize slew time, and subsequently reduces the number of exposures and overall survey depth.
The tradeoff essentially comes down to the relationship between mean pixel loss
reduction and survey depth reduction. If the amount of pixels saved outweighs the amount of
survey depth lost, then the algorithm is valuable to be added to the scheduler. The tradeoff
relationship is evaluated in Figure \ref{fig-trade-off}. We found that, for all three constellations, the tradeoff
between reduction of pixel loss and loss in coadded image depth is so significant that satellite
dodging is overwhelmingly not worth it, provided the simulated constellations we selected are
representative of reality. Since each pointing contains such a large number of pixels, the number
of pixels destroyed by satellite streaks constitutes a very small percentage of the overall image.
Therefore, while the weighted dodging algorithm is effective at reducing satellite streaks in
images, at the current stage, satellite avoidance is mostly unnecessary considering the trade-off
between pixel loss and loss in overall survey depth.

\section{Future Work}
While the satellite dodging algorithm is mostly unnecessary for the three satellite
constellations we considered, many astronomers estimate that there will be sharp increases in
satellite population in the next 5–10 years, which notably overlaps the LSST operations period
(2024–2035). With the dramatic increase in satellite population, the dodging algorithm might
become more relevant. In addition, satellites from other operators may be significantly brighter
than present-day Starlink and OneWeb satellites, and may saturate the LSST Camera’s detectors.
In this scenario, dodging may be required to salvage some science, even if it means losing out on
survey depth.
Related to this, one future work direction involves adding in brightness weighting to the
dodging algorithm. The idea is to only avoid satellites brighter than a certain brightness
threshold. This could potentially reduce the region of avoidance, therefore reducing the loss in
coadded depth — it could reduce the tradeoff between pixel loss and loss in survey depth,
making this preliminary work directly relevant for a successful LSST. Finally, it may be possible
to compute optimal starting locations for observation schedules based on predicted satellite maps
in order to optimize satellite avoidance.

\software{Astropy \citep{astropy2013, astropy2018}, 
          Healpy and HEALpix\footnote{\url{http://healpix.sourceforge.net}} \citep{healpix2005, Zonca2019},
          Matplotlib \citep{Hunter:2007},
          Numpy \citep{harris2020array},
          rubin\_sim \citep{yoachim2022},
          Scipy \citep{2020SciPy-NMeth},
          Shapely\footnote{https://github.com/shapely/shapely}}
          % todo: add citation to newest 2022 astropy paper
